NumPy---Array Programming in Python

### 2020-01-27 body.tex
Edit beginning of ecosystem section

SciPy and Matplotlib are tightly integrated with NumPy.
% to expand the suite of standard algorithms and
% provide basic plotting capabilities.
SciPy provides fundamental algorithms for scientific computing
including mathematical, scientific, and engineering routines.
Matplotlib generates publication-ready figures and visualizations.
The combination of NumPy, SciPy, and Matplotlib with 
an advanced interactive environment like IPython \cite{perez2007ipython}
or Jupyter \cite{Kluyver:2016aa}
provides a complete foundation for array-based scientific computing
and makes interactive data exploration fluid and simple.
The scientific Python ecosystem (Fig.~\ref{fig:ecosystem}) builds on top of
this foundation providing many, widely used technique-specific
libraries\cite{pedregosa2011scikit,vanderwalt2014scikit,SciPyProceedings_11}
and several domain-specific projects and libraries
\cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,2018EGUGA..2012146H}.
NumPy plays a coordinating role for the ecosystem of array-aware
libraries including setting documentation standards,
providing array testing infrastracture,
and coordinating minimum version depedencies.

Exposing this foundation for array programming, as well as the surrounding ecosystem of
tools, in IPython or Jupyter, creates an ideal environment for
interactive, exploratory data analysis where users iteratively inspect,
manipulate, and visualize their data.
Working interactively with their data, users discover through experimentation
bits of code that they stitch together into imperative or functional programs
or narrative computational notebooks.
This rich and productive environment has made the scientific Python ecosystem popular for scientific
research.
As a result, important scientific software packages build on the ecosystem
further exanding it for specific research projects.

@@

Similar examples can be found in physics, earth sciences, biology,
and more.
These researchers depend on NumPy being correct, reliable,
and repeatable.
The rich ecosystem of tools building on NumPy and the larger community
of scientists and programmers using those tools mean that NumPy has a
responsibility to maintain considerable stability.
This conflicts with the need to innovate;
resolving it requires balancing trade-offs.
% our focus is ...
This need creates interesting challenges for NumPy development as the
computational environment scientists are analyzing their data on continues
to evolve and change.
% this leads to a proliferation of arrays

@@

% pandas, xarray, ...
% distributed arrays, delayed execution arrays
% GPU arrays, tensor computation libraries
% sparse arrays
% foreign language arrays

>>>>

SciPy and Matplotlib are tightly coupled with NumPy---in terms of
history, development, and usage.
% to expand the suite of standard algorithms and
% provide basic plotting capabilities.
SciPy provides fundamental algorithms for scientific computing
including mathematical, scientific, and engineering routines.
Matplotlib generates publication-ready figures and visualizations.
The combination of NumPy, SciPy, and Matplotlib together with
an advanced interactive environment like IPython \cite{perez2007ipython}
or Jupyter \cite{Kluyver:2016aa}
provides a complete foundation for array-based scientific computing.
The scientific Python ecosystem (Fig.~\ref{fig:ecosystem}) builds on top of
this foundation to provide several, widely used \emph{technique-specific}
libraries\cite{pedregosa2011scikit,vanderwalt2014scikit,SciPyProceedings_11},
that in turn underlay numerous \emph{domain-specific} projects
\cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,2018EGUGA..2012146H}.
NumPy, as the focal point for the ecosystem of array-aware libraries,
sets documentation standards, provides array testing infrastracture,
and defines minimum version depedencies.

The interactive environment created by this array programming
foundation---along with the surrounding ecosystem of tools---inside of
IPython or Jupyter, is ideally suited to exploratory data analysis.
Users fluidly inspect, manipulate, and visualize their data, and
rapidly iterate to refine programming statements. These statements are
then stitched together into imperative or functional programs, or
narrative computational notebooks.  This rich and productive
environment has made the scientific Python ecosystem hugely popular
for scientific research.

Despite its tight integration, the ecosystem was conceived to have
clear functional separation, with well defined interfaces.  This makes
possible the design of large, complex scientific libraries, that
add specialized functionality for solving specific problems.

%As a result, important scientific software
%packages build on the ecosystem, and further expand it for specific
%research projects.

@@

Similar examples can be found in physics, earth sciences, biology,
and more.
Researchers in these fields depend on NumPy being correct, reliable,
and repeatable.
The rich ecosystem of tools building on NumPy and the larger community
of scientists and programmers using those tools mean that NumPy has a
responsibility to maintain considerable stability.
At times, this conflicts with the need to innovate;
resolving it requires balancing trade-offs.
% our focus is ...
% NumPy purpose engages with this problem by ...
% [this also defines the scope of NumPy]
This need creates interesting challenges for NumPy development as the
computational environment scientists are analyzing their data on continues
to evolve and change.
% this leads to a proliferation of arrays

@@

% pandas, xarray, ...
% numba?
% distributed arrays, delayed execution arrays
% GPU arrays, tensor computation libraries
% sparse arrays
% foreign language arrays
